The impact of forest management on the temperature sensitivity of organic matter turnover

### 2024-11-19 Manuscript_draft/Draft.Rmd
Small changes in discussion

### Role of temperature and moisture in decomposition (model description)
Our model showed that soil temperature and moisture were able to explained more than 50% of variance in soil respiration. Because phenologial patterns in deciduous species can affect sensitivity of soil respiration to temperature (Curiel-Yuste et al 2004), in this model we have also included a seasonality scaling. The seasonal scaling although apparently differentiated slightly between sites, it always presented very low amplitude posteriors, weighting on average only a 5% of the total measured respiration. It was more important in France (between 4.5% and 12.2%) and much less in Spain and Romania (between 2.0% and 3.2%). This is probably related to higher photosynthetic activity variation of Quercus pubescens growing in a Mediterranean climate in France. This would indicate higher changes in root respiration and proliferation that would in turn affect microbial communities. Because in the three study sites there were deciduous Quercus species, another explanation is that soil respiration could have been also affected by the deciduousness of the understorey vegetation (Curiel-Yuste 2004).

## Forest management effect on activation energy and sensitivity to temperature
Generally, our results confirmed the hypothesis 1, since slash addition increased the respiration rates compared to the treatments where the slash was removed. Addition of slash created an artificial litter layer that buffered the effects over soil microclimatic conditions, and provided new habitat and food availability for soil organisms (ref). These results indicate that increases in heterotrophic respiration due to an increment in organic matter availability by slash addition compensated for the decreases in autotrophic respiration and root-associated microbiota due to tree logging and soil disturbance (Borkhuu et al. 2015). Moreover, the strength of the response was also dependent on the forest, being more evident in Spain and France where the tree density was much higher than Romania. In Romania such a low density of trees together with a dense understory of grasses in all treatments might explain the lack of response and indicate high degree of resilience to perturbations.

Our results also confirmed the hypothesis 2, since addition of slash decreased the activation energy of SOM decomposition. We argue that this result is probably related to a reduction in chemical recalcitrance, supporting the carbon quality-temperature hypothesis stating that the biochemically recalcitrant C pool is more sensitive to changes in temperature than the labile C pool (Bossata and Agren 1999). Moreover, the activation energy correlated positively with the Q10, agreeing with Davidson & Janssens, 2006 and Sierra, 2012 that the higher the activation energy the higher is the relative temperature sensitivity. We have also observed a general negative trend of activation energy with time in all management disturbance treatments and all sites. This result also gives support to the C quality temperature hypothesis since fresh inputs from the new vegetation low down the temperature sensitivity of C mineralization as these inputs are more labile to decomposed (Lefevre et al 2014). Still, the growth of new vegetation, within three years from disturbance, did not compensate for the decreases in activation energies in the plots where removal of slash had an effect. Additionally, we cannot rule out the fact that other mechanisms could have contributed to low down activation energies with time such as differences between years in primary production (i.e., contribution of autotrophic respiration) or changes in substrate and enzyme diffusivity driven by differences in soil moisture between years (Davinson and Jansen 2022). In line with this, for example, in the control plots from Spain the activation energy decreased from year zero (2021) to year two (2023) and then it increased a bit, while in disturbed plots it was always descendant. Since changes in activation energy due to the growth of new vegetation is not expected in undisturbed control plots and given that 2022 and 2023 were extremely dry in Spain, probably drought exerted a constrained access of enzymes to substrates (Davinson and Jansen 2022). This suggest that drought effect may have similar impacts on temperature sensitivity as chemical recalcitrance (Levegre et al 2014).

@@

Our analyses revealed the three variables that explained most of the variation of Q10 were related to i) the quality/quantity of substrate (i.e., nitrogen content), ii) the microorganisms, in particular, fungal to bacteria ratio and iii) the soil temperature. Nitrogen fertilization experiments have shown that nitrogen availability decreases the Q10 (Mo et al 2008; Zhang et al 2019), probably related to changes in microbial carbon use efficiency (Manzoni et al 2012). In our experiment, it seems that addition of slash have stimulated decomposition of labile C which increased N availability (Ghosh et al., 2019; Shahbaz et al.,2017), and thus decreased the temperature sensitivity. Our results indicate that SOM decomposition not only depends on the substrate biochemistry but also on the ability of the microbial community to decompose the available substrate. This is in agreement with a recent global-scale study stating that soil microorganism are the main players governing the responses of soil heterotrophic respiration to temperature, and secondly substrate quantity and biochemical recalcitrance (Sáez-Sandino et al 2023).

We did not measure chemical recalcitrance per se, but our results suggest that microbial communities might have a key role in determining Q10 in forest ecosystems. Lower Q10 values associated to the addition of fresh plant material as slash was related to higher fungal to bateria rates. Although both fungi and bacteria have the capacity to decompose litter components (Stursova et al. 2012), fungi has a major role in the decomposition of plant biomass due to their mycelial growth, the ability to translocate nutrients and the possession of a strong enzymatic capacity (Schneider et al. 2012, Tlaskal et al 2016). In contrast, bacteria seems to preferentially gain carbon from live or death fungal biomass and increase at later stages of litter decomposition (Tlaskal et al 2016, Lindahl et al 2010). Since ectomycorrhizal fungal mycelium comprise most of the fungal biomass in forests (Wallander et al 2001), a strong reduction in ectomycorrhizal fungal biomass is expected in all the tree removal disturbance treatments, thus decreasing the fungal to bacteria ratio. As mentioned, this decrease in ectomycorrhizal fungal biomass might be compensated by a rapid growth of saprotrophic fungal species where slash is added. Community data would give more insights about that. Surprinsingly, hydrolitic enzymes were not the main variables affecting Q10. This effect might be masked by the positive correlation of fungal and bacteria biomass with enzymes. Still, most enzymes correlated negatively with Q10. The exception was beta-xylosidase, that correlated positively with Q10 and explained a high proportion of variance. The only microbial group that strongly correlated with beta-xylosidase was the actinobacteria and problably the relationship is determined by the exceptionally high activity of these enzyme in Romania, compared with other sites. This result is consistent with previous studies showing that members of the phylum Actinobacteria are oligotrophic micoorganisms with the ability to compete for resources in a resource-limited environment (Trivedi et al 2013; Bao et al 2021).

>>>>

### Role of temperature and moisture in decomposition (model description)
Our model showed that soil temperature and moisture were able to explained more than 50% of variance in soil respiration. MORE REFS RELATED TO MODELS (Nissan et al 2023,etc.). Because phenologial patterns in deciduous species can affect sensitivity of soil respiration to temperature (Curiel-Yuste et al 2004), in this model we have also included a seasonality scaling. The seasonal scaling although apparently differentiated slightly between sites, it always presented very low amplitude posteriors, weighting on average only a 5% of the total measured respiration. It was more important in France (between 4.5% and 12.2%) and much less in Spain and Romania (between 2.0% and 3.2%). This is probably related to higher photosynthetic activity variation of Quercus pubescens growing in a Mediterranean climate in France. This would indicate higher changes in root respiration and proliferation that would in turn affect microbial communities. Because in the three study sites there were deciduous Quercus species, another explanation is that soil respiration could have been also affected by the deciduousness of the understorey vegetation (Curiel-Yuste 2004).

## Forest management effect on activation energy and sensitivity to temperature
Generally, our results confirmed the hypothesis 1, since slash addition increased the respiration rates compared to the treatments where the slash was removed. Addition of slash created an artificial litter layer that buffered the effects over soil microclimatic conditions, and provided new habitat and food availability for soil organisms (Pérez-Izquierdo et al. in prep). These results indicate that increases in heterotrophic respiration due to an increment in organic matter availability by slash addition compensated for the decreases in autotrophic respiration and root-associated microbiota due to tree logging and soil disturbance (Borkhuu et al. 2015). Moreover, the strength of the response was also dependent on the forest, being more evident in Spain and France where the tree density was much higher than Romania. In Romania such a low density of trees together with a dense understory of grasses in all treatments might explain the lack of response and indicate high degree of resilience to perturbations.

Our results also confirmed the hypothesis 2, since addition of slash decreased the activation energy of SOM decomposition. We argue that this result is probably related to a reduction in chemical recalcitrance, supporting the carbon quality-temperature hypothesis stating that the biochemically recalcitrant C pool is more sensitive to changes in temperature than the labile C pool (Bossata and Agren 1999). Moreover, the activation energy correlated positively with the Q10, agreeing with Davidson & Janssens (2006) and Sierra (2012) that the higher the activation energy the higher is the relative temperature sensitivity. We have also observed a general negative trend of activation energy with time in all management disturbance treatments and all sites. This result also gives support to the C quality temperature hypothesis since fresh inputs from the new vegetation low down the temperature sensitivity of C mineralization as these inputs are more labile to decomposed (Lefevre et al 2014). Still, the growth of new vegetation, within three years from disturbance, did not compensate for the decreases in activation energies in the plots where removal of slash had an effect. Additionally, we cannot rule out the fact that other mechanisms could have contributed to low down activation energies with time such as differences between years in primary production (i.e., contribution of autotrophic respiration) or changes in substrate and enzyme diffusivity driven by differences in soil moisture between years (Davinson and Jansen 2022). In line with this, for example, in the control plots from Spain the activation energy decreased from year zero (2021) to year two (2023) and then it increased a bit, while in disturbed plots it was always descendant. Since changes in activation energy due to the growth of new vegetation is not expected in undisturbed control plots and given that 2022 and 2023 were extremely dry in Spain, probably drought exerted a constrained access of enzymes to substrates (Davinson and Jansen 2022). This suggest that drought effect may have similar impacts on temperature sensitivity as chemical recalcitrance (Levegre et al 2014).

@@

Our analyses revealed the three variables that explained most of the variation of Q10 were related to i) the quality/quantity of substrate (i.e., nitrogen content), ii) the microorganisms, in particular, fungal to bacteria ratio and iii) the soil temperature. Nitrogen fertilization experiments have shown that nitrogen availability decreases the Q10 (Mo et al 2008; Zhang et al 2019), probably related to changes in microbial carbon use efficiency (Manzoni et al 2012). In our experiment, it seems that addition of slash have stimulated decomposition of labile C which increased N availability (Ghosh et al., 2019; Shahbaz et al.,2017), and thus decreased the temperature sensitivity. We did not measure chemical recalcitrance per se, but our results suggest that SOM decomposition not only depends on the substrate biochemistry but also on the ability of the microbial community to decompose the available substrate, having a key role in determining Q10 in forest ecosystems. This is in agreement with a recent global-scale study stating that soil microorganism are the main players governing the responses of soil heterotrophic respiration to temperature, and secondly substrate quantity and biochemical recalcitrance (Sáez-Sandino et al 2023).

Lower Q10 values associated to the addition of fresh plant material as slash was related to higher fungal to bateria rates. Although both fungi and bacteria have the capacity to decompose litter components (Stursova et al. 2012), fungi has a major role in the decomposition of plant biomass due to their mycelial growth, the ability to translocate nutrients and the possession of a strong enzymatic capacity (Schneider et al. 2012, Tlaskal et al 2016). In contrast, bacteria seems to preferentially gain carbon from live or death fungal biomass and increase at later stages of litter decomposition (Tlaskal et al 2016, Lindahl et al 2010). Since ectomycorrhizal fungal mycelium comprise most of the fungal biomass in forests (Wallander et al 2001), a strong reduction in ectomycorrhizal fungal biomass is expected in all the tree removal disturbance treatments, thus decreasing the fungal to bacteria ratio. As mentioned, this decrease in ectomycorrhizal fungal biomass might be compensated by a rapid growth of saprotrophic fungal species where slash is added. Community data would give more insights about that. Surprisingly, hydrolitic enzymes were not the main variables affecting Q10. This effect might be masked by the positive correlation of fungal and bacteria biomass with enzymes. Still, most enzymes correlated negatively with Q10. The exception was beta-xylosidase, that correlated positively with Q10 and explained a high proportion of variance. The only microbial group that strongly correlated with beta-xylosidase was the actinobacteria and problably the relationship is determined by the exceptionally high activity of these enzyme in Romania, compared with other sites. This result is consistent with previous studies showing that members of the phylum Actinobacteria are oligotrophic micoorganisms with the ability to compete for resources in a resource-limited environment (Trivedi et al 2013; Bao et al 2021).
